--- 
title: "Bioprocess Engineering"
shorttitle: "Bioprocess Engineering"
author: "R. Clay Wright"
correspondence: "Clay Wright (<wrightrc@vt.edu>)"
date: "`r Sys.Date()`"

site: bookdown::bookdown_site

params: 
  key: TRUE
  code: TRUE
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = params$code, tidy = TRUE, tidy.opts = list(width.cutoff=50))

options(width = 50)

library(tidyverse)
library(magrittr)
library(ggthemes)
theme_set(theme_few())

#Setup a ???Questions??? answer key in a separate document, per Paul Rougieux https://stackoverflow.com/questions/42960429/excluding-part-of-the-r-markdown-html-notebook-from-the-final-html-output 
#Or in the same document per Yihui Xie! https://stackoverflow.com/questions/32944715/conditionally-display-block-of-markdown-text-using-knitr?noredirect=1&lq=1

library(knitr)
knit_engines$set(answer = function(options) {
  if (options$include && options$echo && options$eval) knit_child(text = options$code)
})

```

<!--- Question Template
#### ???Questions??? 

```{answer, include=params$key}
>* *
``` 

*** 

--->

```{r, eval = FALSE, include = FALSE}
library(bookdown)
bookdown::render_book(input = "index.Rmd")
```

# Introduction

This book has been developed as a companion for the course in Bioprocess Engineering for Biological Systems Engineers at Virginia Tech, BSE3534. It is based on many different resources from the chemical and biological engineering fields (see Additional Resources below) and aims to develop many introductory ideas in kinetics and reactor design through a primarily biological lens. 

## Description and Objectives

This text and course introduce engineering concepts for biological conversion of raw materials to food, pharmaceuticals, fuels, and chemicals. These concepts include enzyme kinetics, metabolic and cell growth kinetics, and design and analysis of reactors, bioreactors, and fermenters. 

### Course Objectives

After reading this text and working associated examples and assignments, students will be able to: 

- Model chemical and biochemical reactions using mass balances and kinetic rate equations. 
- Analyze and formulate mechanisms and rate equations for enzymatic reactions. 
- Estimate kinetic parameters from experimental data.
 -Understand soluble and immobilized enzyme technologies for the production of industrial and medical products.
- Model microbial growth and metabolism.
- Identify and design bioreactors for various products and organisms.
- Specify required practices and technologies to safely and effectively utilize genetically engineered organisms for biomanufacturing.


## Additional Resources

Chemical Reactions and Chemical Reactors  
by George Willard Roberts  
Find a copy: http://www.worldcat.org/oclc/441742755

Elements of Chemical Reaction Engineering, 6th Edition  
by H Fogler  
Available online: https://www.worldcat.org/oclc/1192531500 

Fundamentals of Modern Bioprocessing 1st Edition  
by Sarfaraz K. Niazi, Justin L. Brown   
FREE PDF download: https://www.worldcat.org/oclc/958799491  

Bioprocess Engineering: Basic Concepts, 3rd Edition  
by Shuler, Kargi & DeLisa  
Available online: https://www.worldcat.org/oclc/981256254  

Integrated Bioprocess Engineering  
by Clemens Posten  
Available online: https://www.worldcat.org/oclc/1032686988  

Bioprocess Engineering  
by Shijie Liu  
Available online: http://www.worldcat.org/oclc/1173991723

## Welcome to Bioprocess Engineering (BSE 3534) - Syllabus Rundown

### Course Philosophy

This course emphasizes interactive, problem-focused learning. Through group work, in-class problem-solving, and step-by-step examples, you'll develop the skills to approach and solve engineering challenges. Participation, whether live or asynchronous, is key to success.

### Community Guidelines

Our collaborative environment values respect and inclusivity. Actively listen, include all voices, and constructively engage with different perspectives. If issues arise, contact the instructor or TAs.

### Tools and Platforms

- **Canvas**: All assignments will be submitted through Canvas. We will also keep links to course materials in a Google Drive folder updated on Canvas, but do let us know if any links are broken.
- **Google Drive**: Shared documents for live collaboration. Ask questions directly in documents as comments. Feel free to suggest changes too!


### Assessments
- Formative (skill-building) assessments include collaborative homework and active learning credits.
- Summative assessments (skill-evaluating) will take the form of 3 exams. 

### Modules Overview
1. Modeling chemical reactions: Kinetics and reactors
2. Biochemical/Enzyme kinetics  
3. Cell growth kinetics and reactor design

Let’s build a supportive and engaging learning community. Reach out anytime via Piazza or email with questions or suggestions. Looking forward to a great semester!


